Starcode: an exact algorithm for sequence clustering

### 2014-06-19 doc/document.tex
Corrected some more typos in the manuscript.

\subsection{Parallelization}
To parallelize the search, queries are separated into contiguous blocks after
sorting. The matching algorithm proceeds in two phases. During the first phase,
an independent trie is created for each block and filled with its sequence using
the seek and construct process described above. In the second
phase, blocks are only queried against but not inserted into the tries built in
the first phase. If the
queries are segregated into $N$ blocks, the first phase consists of $N$ seek and
construct jobs, and the second consists of $N(N-1)/2$ query jobs. Since the jobs
show little dependence on each other, the matching algorithm can be efficiently
parallelized provided $N$ is larger than the number of independent threads.

@@

In message passing clustering, read counts are distributed equally among the
closest neighbors of each barcode if they are at least 5 times more frequent.
The remaining sequences at the end of the process are considered
canonical, and their associated count is the estimated cluster size. A barcode
is assigned to a cluster if all its read count are eventually given to the
corresponding canonical barcode. The barcodes for which the read counts are
split between different canonical barcodes are not assigned to any cluster. The
reason for imposing a factor 5 or larger in order to transfer the read counts
is that barcodes with similar frequencies are not likely derived from each other
through sequencing errors. More likely they are either unrelated, or they both
derive from another more abundant barcode.

@@

Denoted as \emph{sequential R} is the sequential algorithm of
\cite{pmid23953119} which implemented in R and relied on the \texttt{stringdist}
function of the \texttt{stringdist} package.

@@

We generated 1 million random 40-mers that were duplicated 47 times. For each
40-mer, we inserted 3 addional mutant sequences to the pool by sampling again
3 nucleotides
at random positions. The probability that such a mutant sequence is within a
distance 3 of any other 40-mer is of the order of 10-12 and can be neglected.
Each cluster thus consists of 50 sequences, of which 3 have a distance at most
3 from the canonical representant.

Using 20 cores, starcode clustered the shuffled 50 million sequences in 4 minutes
without a single error (the output consisted of 1 million clusters of size 50).
We submitted the unique sequences of the dataset to CD-HIT and USEARCH. The
results are summarized in Figure~\ref{fig:04}. The running times of CD-HIT and USEARCH
were 59 and 64 minutes respectively, and they identified 2,362,352 and 2,739,954
clusters respectively. Both algorithms rely on heuristics, fact that explains why the
number of clusters is off target. These results clearly demonstrate that starcode
outperforms other clustering algorithms in speed and accuracy on this particular task.

@@

For this test, we used the same dataset described above, in which every
sequence has at least 2 distinct matches at a distance 3 (mutant sequences
from the same cluster may be at a distance up to 6). This means that we
can use matches with a single hit (themselves) as a lower bound on the inaccuracy.
We used unique sequences to build an index and queried the index with the same
sequences. Figure~\ref{fig:05} shows the running time of starcode along with the
decomposition of the running time for the three mappers. Note that time to
extract unique sequences is always the same because we used the Linux
\texttt{sort -u} command in every case. It appears that the running time of
starcode is about the same as the time required by BWA (the fastest aligner in
this test) to index and run the query proper. Figure~\ref{fig:05} shows the
lower bound on
the error rate, as estimated by the proportion of single hits. BWA was found to
be the most accurate aligner with a lower bound of 19\% errors. This measure
is indeed an underestimate of the true error rate, but given that starcode makes no
mistake on this dataset, a more accurate measure is not necessary. In conclusion,
starcode runs slightly faster than short read mappers when all steps are taken
into account, and clearly outperforms them in precision on this particular problem.

\subsection{Clustering TRIP barcodes}
In the course of setting up the TRIP technology in our lab \citep{pmid23953119}, we
realized the need to develop efficient algorithms to cluster similar sequences.
Briefly, the principle of TRIP (Thousands of Reporters Integrated in Parallel)
is to tag reporter transcripts with random barcodes and measure the abundance
of barcodes in the RNA as a proxy for gene expression. There is no reference to
match aberrant barcodes against, because the tagging sequences are unknown.
Instead, barcodes are matched against each other and clustered by similarity to
infer canonical sequences.

We tested the efficiency of starcode on the TRIP dataset from
\cite{pmid23953119}. In the experiment labelled mPGKA, we identified
24.1 million (91\%) barcode-containing reads out of 26.6 million, consisting
of approximately 223,000 and 220,000 unique barcode sequences for PCR replicates
1 and 2 respectively. Following the authors, we kept only the barcodes with
at least 5 counts and performed clustering as described: ``First we sorted
barcodes according to their counts. Then, for each barcode (starting from the
most frequent one), we identified and removed all its mutant versions, defined
as barcodes within a Hamming distance of 2.'' This method is here on referred
to as the ``sequential algorithm''. When replacing the Hamming distance by the
Levenshtein distance, starcode produced exactly the same output as the sequential
algorithm. The running time of starcode averaged over the replicates was
2.90 seconds, \textit{versus} more than 3 minutes 40 seconds, which represents a
75-fold speedup.

The performance of the sequential algorithm relies on the arbitrary exclusion
of reads with less than 5 counts, the main purpose of which is to reduce the
computational burden. When all barcodes were kept, the average running time of
starcode was 10.27 seconds, \textit{versus} $>$ 6 hours for the sequential
algorithm. Using the Hamming distance, as the authors originally suggested,
decreased the average running time of the sequential algorithm to about 3 hours
and 30 minutes (and
the output differed from that of starcode). Note that in all the cases
mentioned above, starcode was run with a single core to compare the algorithms
based on similar computer resources. In conclusion, the barcode clustering
problem can be simplified by various heuristics, but starcode brings down the
running time to nearly instantaneous, and thereby obviates the need for such
arbitrary heuristics.

@@

We set up a test based on the meningitis-causing agent
\textit{Neisseria meningitidis}. The genome of this bacterium is interspersed
with a frequent 12 bp sequence known as DNA uptake sequence \citep{pmid10673000}. We
extracted the 12-mers from both orientations of  the 2.19 Mb genome, yielding
4.39 million 12-mers, consisting of 2.77 million unique sequences. Clustering
the 12-mers with starcode within a Levenshtein distance of 2 took less
than 3 minutes with 20 cores. We identified the known DNA uptake sequence of
Neisseria meningitidis (ATGCCGTCTGAA) as the most abundant 12-mer, comprising
1,466 exact matches and 4,625 inexact matches. This result testifies to the
fact that starcode can be used to identify biologically relevant motifs in
bacterial genomes.

To test starcode on the scale of a metazoan genome, we used a Drosophila
dataset from \cite{pmid20888037}. The study describes two
signatures of chromatin proteins called Red and Yellow, both corresponding
to transcribed regions of the genome. The distinction is based on epigenomic
features only, which prompted us to identify sequence features that could
discriminate Red \textit{versus} Yellow regions. To this end, we used starcode to
cluster 18-mers from both regions.

The total coverage of Red and Yellow chromatin in the Drosophila genome is
10.7 Mb and 21.3 Mb respectively, corresponding to 21.5 million and 42.5 million
18-mers, respectively. We used starcode to identify clusters of 18-mers within
a Levenshtein distance of 3 (the running time was approximately 35 minutes for
Red 18-mers and 70 minutes for Yellow 18-mers using 20 threads). The frequencies
of the most abundant 18-mers showed that Red chromatin has a lower sequence
complexity than Yellow chromatin (not shown). The most discriminating 18-mer
was (GA)$_9$, which showed a 4.8 fold enrichment in Red \textit{versus} Yellow
chromatin.
We counted 832 exact matches and 6,749 inexact matches over 3,880 distinct
18-mers in Red chromatin, showing that a larger portion of the signal comes
from inexact matches. Taken together, these results show that starcode can be
used to cluster short k-mers from metazoan-scale datasets.

\section{Discussion and conclusion}
Through the parallel poucet search algorithm, starcode implements an exact
sequence clustering algorithm that is even faster than widely used heuristics.
By design, starcode is tailored to process high throughput sequencing data on
multi-core platforms. Our benchmark shows that starcode is superior in speed
and accuracy compared to other methods on the problem of clustering short random
sequences. Part of the reason is that such task is stretching the tested
software far from their initial design. Sequence clustering algorithms are
meant to to cluster long biological sequences, while short read mappers are
meant to map reads on potentially large genomes. In this respect, starcode
fills a need arising from the development of barcoding technologies.

The speed of starcode also makes it proficient at more general clustering
problems, such as identifying enriched k-mers in genomes and in experimental
data. Here we have given two examples of such applications. In the first, we
recover a known enriched 12-mer in the genome of \textit{Neisseria meningitidis}.
In the second, we identify a sequence feature that discriminates genomic
regions found in two different chromatin types in Drosophila. GA repeats are
known to be bound in vivo by the protein GAF \citep{pmid12601174}, which suggests
that this transcription factor is instrumental in setting the distinction
between Red and Yellow chromatin.

The current version of starcode has been mildly optimized for memory
consumption. The memory footprint depends on the number of sequences to
cluster (because the sequences of the input set are loaded in memory as a
group of tries) and on the mean number of matches per sequence. Every match
has to be stored until the clustering phase and may represent a heavy
memory overhead. As counterintuitive as it may seem, long queries will usually
impose a lower memory footprint because the matches between sequences are less
frequent.

The speed of starcode could also benefit other applications, such as searching
shared subsequences in genome assembly. However, starcode supports only global
and not local alignment. A workaround is to compare k-mers as we have done
in the examples above, but this approach is expected to be memory demanding.
Even though extending the starcode algorithm to support clustering of protein
sequences is straightforward, it is not expected to meet the performance level
of the presented algorithm. On the one hand, there are more letters in the protein
alphabet, so the shared prefix between consecutive queries is inherently
shorter on average, which defeats the poucet search. On the other hand, if
a threshold distance of the order of 1 per 10 nucleotides is sufficient to
identify similar DNA sequences, a distance of 1 per 10 residues is
unrealistically low for protein sequences, and this parameter critically
determines the speed of starcode. More generally, storing computational
intermediates for shared prefixes could find some applications in other
algorithms such as short read mappers. The idea of the poucet search seems
simple in retrospect, but it is a powerful way to tap into the data
structuration provided by string sorting.

>>>>

\subsection{Parallelization}
To parallelize the search, queries are separated into contiguous blocks after
sorting. The matching algorithm proceeds in two phases. During the first phase,
an independent trie is created and filled with the sequences
of one block using the seek and construct process described above. In the second
phase, blocks of sequences are queried against but not inserted into the tries
built in the first phase. If the
queries are segregated into $N$ blocks, the first phase consists of $N$ seek and
construct jobs, and the second consists of $N(N-1)/2$ query jobs. Since the jobs
show little dependence on each other, the matching algorithm can be efficiently
parallelized provided $N$ is larger than the number of independent threads.

@@

In message passing clustering, read counts are distributed equally among the
closest neighbors of each barcode if they are at least 5 times more frequent.
The remaining sequences at the end of the process are considered
canonical, and their associated count is the estimated cluster size. A barcode
is assigned to a cluster if all its read counts are eventually given to the
corresponding canonical barcode. The barcodes for which the read counts are
split between different canonical barcodes are not assigned to any cluster. The
reason for imposing a factor 5 or larger in order to transfer the read counts
is that barcodes with similar frequencies are not likely derived from each other
through sequencing errors. More likely they are either unrelated, or they both
derive from another more abundant barcode.

@@

Denoted as \emph{sequential R} is the sequential algorithm of
\cite{pmid23953119} which is implemented in R and relied on the \texttt{stringdist}
function of the \texttt{stringdist} package.

@@

We generated 1 million random 40-mers that were duplicated 47 times. For each
40-mer, we inserted 3 addional mutant sequences to the pool by sampling again
3 nucleotides
at random positions. The probability that such a mutant sequence is within a
distance 3 of any other 40-mer is of the order of $10^{-12}$ and can be neglected.
Each cluster thus consists of 50 sequences, of which 3 have a distance at most
3 from the canonical representant.

Using 20 cores, starcode clustered the shuffled 50 million sequences in 4 minutes
without a single error (the output consisted of 1 million clusters of size 50).
We submitted the unique sequences of the dataset to CD-HIT and USEARCH. The
results are summarized in Figure~\ref{fig:04}a and \ref{fig:04}b. The running
times of CD-HIT and USEARCH
were 59 and 64 minutes respectively, and they identified 2,362,352 and 2,739,954
clusters respectively. Both algorithms rely on heuristics, fact that explains why the
number of clusters is off target. These results clearly demonstrate that starcode
outperforms other clustering algorithms in speed and accuracy on this particular task.

@@

For this test, we used the same dataset described above, in which every
sequence has at least 2 distinct matches at a distance 3 (mutant sequences
from the same cluster may be at a distance up to 6). This means that we
can use matches with a single hit (themselves) as a lower bound on the inaccuracy.
We used unique sequences to build an index and queried the index with the same
sequences. Figure~\ref{fig:04}c shows the running time of starcode along with the
decomposition of the running time for the three mappers. Note that the time to
extract unique sequences is always the same because we used the Linux
\texttt{sort -u} command in every case. It appears that the running time of
starcode is about the same as the time required by BWA (the fastest aligner in
this test) to index and run the query proper. Figure~\ref{fig:04}d shows the
lower bound on
the error rate, as estimated by the proportion of single hits. BWA was found to
be the most accurate aligner with a lower bound of 19\% errors. This measure
is indeed an underestimate of the true error rate, but given that starcode makes no
mistake on this dataset, a more accurate measure is not necessary. In conclusion,
starcode runs slightly faster than short read mappers when all steps are taken
into account, and clearly outperforms them in precision on this particular problem.

\subsection{Clustering TRIP barcodes}
In the course of setting up the TRIP technology in our laboratory
\citep{pmid23953119}, we
realized the need to develop efficient algorithms to cluster similar sequences.
Briefly, the principle of TRIP (Thousands of Reporters Integrated in Parallel)
is to tag reporter transcripts with random barcodes and measure the abundance
of barcodes in the RNA as a proxy for gene expression. There is no reference to
match aberrant barcodes against, because the tagging sequences are unknown.
Instead, barcodes are matched against each other and clustered by similarity to
infer canonical sequences.

We tested the efficiency of starcode on the TRIP dataset from
\cite{pmid23953119}. In the experiment labelled mPGKA, we identified
24.1 million (91\%) barcode-containing reads out of 26.6 million, consisting
of approximately 223,000 and 220,000 unique barcode sequences for PCR replicates
1 and 2, respectively. Following the authors, we kept only the barcodes with
at least 5 counts and performed clustering as described: ``First we sorted
barcodes according to their counts. Then, for each barcode (starting from the
most frequent one), we identified and removed all its mutant versions, defined
as barcodes within a Hamming distance of 2.'' This method is here on referred
to as the ``sequential algorithm''. When replacing the Hamming distance by the
Levenshtein distance, starcode produced exactly the same output as the sequential
algorithm. The running time of starcode averaged over the replicates was
2.90 seconds, \textit{versus} more than 3 minutes 40 seconds, which represents a
75-fold speedup.

The performance of the sequential algorithm relies on the arbitrary exclusion
of reads with less than 5 counts, the main purpose of which is to reduce the
computational burden. When all barcodes were kept, the average running time of
starcode was 10.27 seconds, \textit{versus} $>$ 6 hours for the sequential
algorithm. Using the Hamming distance, as the authors originally suggested,
decreased the average running time of the sequential algorithm to about 3 hours
and 30 minutes (and
the output differed from that of starcode). Note that in all the cases
mentioned above, starcode was run with a single core to compare the algorithms
based on similar computer resources. In conclusion, the barcode clustering
problem can be simplified by various tricks, but starcode brings down the
running time to nearly instantaneous, and thereby obviates the need for such
arbitrary heuristics.

@@

We set up a test based on the meningitis-causing agent
\textit{Neisseria meningitidis}. The genome of this bacterium is interspersed
with a frequent 12 bp sequence known as DNA uptake sequence \citep{pmid10673000}. We
extracted the 12-mers from both orientations of  the 2.19 Mb genome, yielding
4.39 million 12-mers, consisting of 2.77 million unique sequences. Clustering
the 12-mers with starcode within a Levenshtein distance of 2 took less
than 3 minutes with 20 cores. We identified the known DNA uptake sequence of
\textit{Neisseria meningitidis} (ATGCCGTCTGAA) as the most abundant 12-mer, comprising
1,466 exact matches and 4,625 inexact matches. This result testifies to the
fact that starcode can be used to identify biologically relevant motifs in
bacterial genomes.

To test starcode on the scale of a metazoan genome, we used a \textit{Drosophila}
dataset from \cite{pmid20888037}. The study describes two
signatures of chromatin proteins called Red and Yellow, both corresponding
to transcribed regions of the genome. The distinction is based on epigenomic
features only, which prompted us to identify sequence features that could
discriminate Red \textit{versus} Yellow regions. To this end, we used starcode to
cluster 18-mers from both regions.

The total coverage of Red and Yellow chromatin in the \textit{Drosophila} genome is
10.7 Mb and 21.3 Mb respectively, corresponding to 21.5 million and 42.5 million
18-mers, respectively. We used starcode to identify clusters of 18-mers within
a Levenshtein distance of 3 (the running time was approximately 35 minutes for
Red and 70 minutes for Yellow chromatin using 20 threads). The most discriminating
18-mer was (GA)$_9$, which showed a 4.8 fold enrichment in Red \textit{versus} Yellow
chromatin.
We counted 832 exact matches and 6,749 inexact matches over 3,880 distinct
18-mers in Red chromatin, showing that a larger portion of the signal comes
from inexact matches. Taken together, these results show that starcode can be
used to cluster short k-mers from metazoan-scale datasets.

\section{Discussion and conclusion}
Through the parallel poucet search algorithm, starcode implements an exact
sequence clustering algorithm that is even faster than widely used heuristics.
By design, starcode is tailored to process high throughput sequencing data on
multi-core platforms. Our benchmark shows that starcode is superior in speed
and accuracy compared to other methods on the problem of clustering short random
sequences. Part of the reason is that such task is stretching the tested
software far from their initial design. Sequence clustering algorithms are
meant to cluster long biological sequences, while short read mappers are
meant to map reads on potentially large genomes. In this respect, starcode
fills a need arising from the development of barcoding technologies.

The speed of starcode also makes it proficient at more general clustering
problems, such as identifying enriched k-mers in genomes and in experimental
data. Here we have given two examples of such applications. In the first, we
recover a known enriched 12-mer in the genome of \textit{Neisseria meningitidis}.
In the second, we identify a sequence feature that discriminates genomic
regions found in two different chromatin types in \textit{Drosophila}. GA
repeats are known to be bound \textit{in vivo} by the protein GAF
\citep{pmid12601174}, which suggests
that this transcription factor is instrumental in setting the distinction
between Red and Yellow chromatin.

The current version of starcode has been mildly optimized for memory
consumption. The memory footprint depends on the number of sequences to
cluster (because the sequences of the input set are loaded in memory as a
group of tries) and on the mean number of matches per sequence. Every match
has to be stored until the clustering phase, which may represent a large
memory overhead. As counterintuitive as it may seem, long queries will usually
impose a lower memory footprint because the matches between sequences are less
frequent.

The speed of starcode could also benefit other applications, such as searching
shared subsequences during genome assembly. However, starcode supports only global
and not local alignment. A workaround is to compare k-mers as we have done
in the examples above, but this approach is expected to be memory demanding.
Even though extending the starcode algorithm to support clustering of protein
sequences is straightforward, it is not expected to meet the performance level
of the presented algorithm. On the one hand, there are more letters in the protein
alphabet, so the shared prefix between consecutive queries is inherently
shorter on average, which defeats the poucet search. On the other hand, if
a threshold distance of the order of 1 per 10 nucleotides is sufficient to
identify similar DNA sequences, a distance of 1 per 10 residues is
unrealistically low for protein sequences, and this parameter critically
determines the speed of starcode. More generally, storing computational
intermediates for shared prefixes could find some applications in other
algorithms such as short read mappers. The idea of the poucet search seems
simple in retrospect, but it is a powerful way to tap into the data
structuration provided by string sorting.
